dispRity metrics

### 2016-05-25 doc/dispRity-metrics.Rmd
Metrics manual update

>>>>

Metrics implemented in `dispRity`
===========================================

Several disparity metrics are directly implemented in the `dispRity` package.
The detailed list can be found in `?dispRity.metric` along with some explanations on each metric.

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `variances` | The variance of each dimension | `dispRity` |
2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `mode.val` | The modal value | `dispRity` |

1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).

2: Note that by default, the centroid is the centroid of the elements.
It can however be fixed to an arbitrary value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.

## Equations and implementations:
Some of the functions described below are directly implemented in the `dispRity` package and do not require any dependencies.

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}
With *k* being the number of dimensions, $\lambda_i$ is the eigen value of each dimension [R implementation line 140](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}
With *k* being the dimensions [R implementation line 220](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}
With *k* being the dimensions [R implementation line 100](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}
With *k* being the dimensions and $\sigma^{2}$ their variance [R implementation line 95](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}
With *n* being each element in the ordinated space, *k* being its dimensions and $Centroid_{k}$ their mean (or can be set to an arbitrary value) and  [R implementation line 111](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).

### Volumes and surface metrics
```{r}
# First lets create a small dummy ordinate matrix
set.seed(1)
dummy_space <- space.maker(10, 5, rnorm)

# Let's then calculate disparity as the ellipsoid volume:
summary(dispRity(dummy_space, metric = ellipse.volume))

# Which is the equivalent of doing (with some rounding):
ellipse.volume(dummy_space)

# Now for the convex hull surface:
summary(dispRity(dummy_space, metric = convhull.surface))

# And for it's volume:
summary(dispRity(dummy_space, metric = convhull.volume))

# Note that the more the dimensions, the more the volume tends towards 0!
summary(dispRity(space.maker(10,9, rnorm), metric = convhull.volume))

# The estimation of the hypervolume is a bit slower:
summary(dispRity(dummy_space, metric = hyper.volume))

# Note that the hypervolume is different of the convex hull volume!
# This function also intakes more arguments that can be fed to dispRity
summary(dispRity(dummy_space, metric = hyper.volume,
                 repsperpoint = 100, quantile = 0.5))
# Which gives quiet different results!
```

### Diagonal, ranges and variances
```{r}
# Or we can use other more straightforward metrics such as the diagonal of our
# dummy space (i.e. the longest euclidean distance in our 5 dimensional space)
summary(dispRity(dummy_space, metric = diagonal))

# Another type of metrics, based on the range are used a lot in palaeobiology.
summary(dispRity(dummy_space, metric = ranges))
# Note that this gives a distribution of metrics now (i.e. the range for our 5
# dimensions).
# This distribution can be summarised using metrics such as the sum or product
# of ranges:
summary(dispRity(dummy_space, metric = c(sum, ranges)))
summary(dispRity(dummy_space, metric = c(prod, ranges)))

# The same can be done with the variances
summary(dispRity(dummy_space, metric = variances))
summary(dispRity(dummy_space, metric = c(sum, variances)))
summary(dispRity(dummy_space, metric = c(prod, variances)))
# Note that neither the sum nor the product of the variances are really
# representative of the variances distribution!
```

### Centroids metrics
```{r}
# The centroid metric allows to calculate the distances between each elements
# and their centroids:
summary(dispRity(dummy_space, metric = centroids))
# Note that this also gives a distribution

# Another interesting use of this function is for looking at both the average
# spread of the elements (or the entire distribution!) and their position within
# the dummy space by looking at the distances between the elements and the
# center of the dummy space (or an arbitrary point)

centroids_to_group <- dispRity(dummy_space, metric = centroids)
centroids_to_space <- dispRity(dummy_space, metric = centroids,
                               centroid = c(rep(0, 5)))

# Average spread
summary(dispRity(centroids_to_group, metric = median))
# Average position within de dummy space
summary(dispRity(centroids_to_space, metric = median))
```
